---
title: "Getting started with ghcm"
output:
    bookdown::html_document2:
      base_format: rmarkdown::html_vignette
      number_sections: false
bibliography: refs.bib
csl: ieee.csl
header-includes:
   - \usepackage{bbm}
vignette: >
  %\VignetteIndexEntry{ghcm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
\newcommand{\independent}{\mbox{${}\perp\mkern-11mu\perp{}$}}
\newcommand{\notindependent}{\mbox{${}\not\!\perp\mkern-11mu\perp{}$}}
\newcommand{\cond}{\,|\,}
\newcommand{\ind}{\mathbf{1}}

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align="center",
  dpi=150,
  fig.width=7,
  fig.height=5,
  out.width="100%"
)
options(rmarkdown.html_vignette.check_title = FALSE) 
```

```{r setup, include=FALSE}
library(ghcm)
```

# Introduction

`ghcm` is an R package used to perform conditional independence tests for densely observed functional data.

This vignette gives a brief overview of the usage of the `ghcm` package. We give a brief presentation of the idea behind the GHCM and the conditions under which the test is valid. Subsequently, we provide several examples of the usage of the `ghcm` package by analysing a simulated data set.

# The Generalised Hilbertian Covariance Measure (GHCM)
In this section we briefly describe the idea behind the GHCM. For the full technical details and theoretical results, see [@Lundborg2021].

Let $X$, $Y$ and $Z$ be random variables of which we are given $n$ i.i.d. observations $(X_1, Y_1, Z_1), \dots, (X_n, Y_n, Z_n)$ and where $X$, $Y$ and $Z$ can be either scalar or functional. Existing methods, such as the GCM [@GCM] implemented in the `GeneralisedCovarianceMeasure` package [@GCM-package], can deal with most cases where both $X$ and $Y$ are scalar hence our primary interest is in the cases where at least one of $X$ and $Y$ are functional.

The GHCM estimates the squared Hilbert-Schmidt norm of the expected conditional covariance of $X$ and $Y$ given $Z$, $\| \mathscr{K} \|_{\mathrm{HS}}^2$, and rejects the hypothesis $X \independent Y \cond Z$ if this quantity is too large. We denote by $\langle f, g \rangle$ the inner product between $f$ and $g$.  We compute the GHCM as follows.

1. Regress $X$ on $Z$ and $Y$ on $Z$ yielding residuals $\hat{\varepsilon}$ and $\hat{\xi}$, respectively.
2. Compute the test statistic
$$
T = \frac{1}{n} \sum_{i=1}^n \sum_{j=1}^n \langle \hat{\varepsilon}_i, \hat{\varepsilon}_j \rangle \langle \hat{\xi}_i, \hat{\xi}_j \rangle.
$$
3. Compute the $n \times n$ matrix $\Gamma$ by setting 
$$
\Gamma_{ij} := \langle \hat{\varepsilon}_i, \hat{\varepsilon}_j \rangle \langle \hat{\xi}_i, \hat{\xi}_j \rangle.
$$
4. Compute the $d \leq n-1$ non-zero eigenvalues $(\lambda_k)_{k=1}^d$ of
$$
C := \frac{1}{n-1} (\Gamma - J \Gamma - \Gamma J + J \Gamma J),
$$
where $J$ is the $n \times n$ matrix with all entries equal to $1/n$.
5. Produce a $p$-value by setting
$$
p = \mathbb{P}\left(\sum_{k=1}^d \lambda_k \zeta_k^2 > T\right),
$$
where $\zeta_1, \dots, \zeta_d$ are i.i.d. standard Gaussian random variables.

Assuming that the regression methods perform sufficiently well, the GHCM has uniformly distributed $p$-values when the null is true. It should be noted that there are situations where $X \notindependent Y \cond Z$ but the GHCM is unable to detect this dependence for any sample size, since $\mathscr{K}$ can be zero in this case.

The GHCM as implemented in this package assumes that the all the residuals of the $X$ on $Z$ regression are evaluated on the same grid and that residuals of the $Y$ on $Z$ regression are evaluated on the same grid. However, these two grids can be distinct. 


# Example applications on a simulated dataset
To give concrete examples of the usage of the package, we perform conditional independence tests on a simulated data set consisting of both functional and scalar variables. The functional variables are observed on a common equidistant grid of $101$ points on $[0, 1]$. 

```{r}
library(ghcm)
set.seed(111)
data(ghcm_sim_data)
grid <- seq(0, 1, length.out=101)
colnames(ghcm_sim_data)
```
`ghcm_sim_data` consists of 500 observations of the scalar variables $Y_1$ and $Y_2$ and the functional variables $X$, $Z$ and $W$. The curves and the mean curve for functional data can be seen in Figures \@ref(fig:plot-X), \@ref(fig:plot-Z) and \@ref(fig:plot-W).

```{r plot-X, fig.cap="Plot of $X$ with the estimated mean curve in red.", echo=FALSE}
matplot(grid, t(ghcm_sim_data$X), type="l",
        col=rgb(0,0,0,0.1) , lty=1, xlab="",
        ylab="")
lines(grid, colMeans(ghcm_sim_data$X), col=2,
      lty=2, lwd=2)
title("X")
```

```{r plot-Z, fig.cap="Plot of $Z$ with the estimated mean curve in red.", echo=FALSE}
matplot(grid, t(ghcm_sim_data$Z), type="l",
        col=rgb(0,0,0,0.1) , lty=1, xlab="",
        ylab="")
lines(grid, colMeans(ghcm_sim_data$Z), col=2,
      lty=2, lwd=2)
title("Z")
```

```{r plot-W, fig.cap="Plot of $W$ with the estimated mean curve in red.", echo=FALSE}
matplot(grid, t(ghcm_sim_data$W), type="l",
        col=rgb(0,0,0,0.1) , lty=1, xlab="",
        ylab="")
lines(grid, colMeans(ghcm_sim_data$W), col=2,
      lty=2, lwd=2)
title("W")
```

In all of the upcoming examples we will use functions from the `refund` R-package [@refund] to perform regressions. We will not attempt to justify the validity of the regression models we use here as the upcoming tests are only included for illustrative purposes. In actual applications of the GHCM however, it is critical that the regression methods employed estimate the conditional expectations $\mathbb{E}(X \cond Z)$ and $\mathbb{E}(Y \cond Z)$ sufficiently well for the $p$-values to be valid. Any use of the GHCM should be prefaced by an analysis of the performance of the regression methods in use.

## Testing independence of scalar variables given functional variables
We first test whether $Y_1$ and $Y_2$ are conditionally independent given the functional variables. This is relevant if, say, we're trying to predict $Y_1$ and we want to know whether including $Y_2$ as a predictor would be helpful. A naive correlation-based approach would suggest that $Y_2$ could be relevant since:
```{r}
cor(ghcm_sim_data$Y_1, ghcm_sim_data$Y_2)
```
To perform the conditional independence test, we need a scalar-on-function regression method and we will use the `pfr` function from the `refund` package [@refund] with `lf`-terms. We run the test in the following code:
```{r scalar-given-functional-test}
library(refund)
m_1 <- pfr(Y_1 ~ lf(X) + lf(Z) + lf(W) , data=ghcm_sim_data)
m_2 <- pfr(Y_2 ~ lf(X) + lf(Z) + lf(W), data=ghcm_sim_data)
test <- ghcm_test(resid(m_1), resid(m_2))
print(test)
```
We get a $p$-value of 0.958. It should be noted that since the asymptotic distribution of the test statistic depends on the underlying distribution, there is no way to know the $p$-value from just the test statistic alone, hence the test_statistic is not reported. 

## Testing independence of a scalar and functional variable given a functional variable
We now test whether $Y_1 \independent X \cond Z$. This is relevant if we're interested in modelling $Y_1$ and want to determine whether $X$ should be included in a model that already includes $Z$. We can plot $X$ and $Z$ and color the curves based on the value of $Y_1$ as can be seen in Figures \@ref(fig:Z-Y-plot) and \@ref(fig:X-Y-plot) below. 

```{r Z-Y-plot, echo=FALSE, fig.cap="Plot of $Z$ with colors based on the value of $Y_1$."}
library(ggplot2)
library(reshape2)

Z_df <- as.data.frame(ghcm_sim_data$Z)
colnames(Z_df) <- grid
Z_df$Y_1<- ghcm_sim_data$Y_1
Z_df$id <- 1:nrow(ghcm_sim_data)
tmp <- melt(Z_df, id.var=c("id", "Y_1"), variable.name="grid")
tmp$grid <- as.numeric(levels(tmp$grid))[tmp$grid]


ggplot(tmp, aes(x=grid, y=value, group=id)) + geom_line(aes(color=Y_1)) +
  theme_bw() + scale_x_continuous(name="") + scale_y_continuous(name="") +
  scale_color_gradient(name=expression(Y[1]))
```

```{r X-Y-plot, echo=FALSE, fig.cap="Plot of $X$ with colors based on the value of $Y_1$."}
library(ggplot2)
library(reshape2)

X_df <- as.data.frame(ghcm_sim_data$X)
colnames(X_df) <- grid
X_df$Y_1<- ghcm_sim_data$Y_1
X_df$id <- 1:nrow(ghcm_sim_data)
tmp <- melt(X_df, id.var=c("id", "Y_1"), variable.name="grid")
tmp$grid <- as.numeric(levels(tmp$grid))[tmp$grid]


ggplot(tmp, aes(x=grid, y=value, group=id)) + geom_line(aes(color=Y_1)) +
  theme_bw() + scale_x_continuous(name="") + scale_y_continuous(name="") +
  scale_color_gradient(name=expression(Y[1]))
```
It appears that both of the functional variables contain information about $Y_1$. To use the GHCM for this test, in addition to the scalar-on-function regression employed in the previous section, we will need to be able to perform function-on-function regressions. This is done using the `pffr` function in the `refund` package [@refund] with `ff` terms. We run the test in the following code:
```{r Y-X-Z-test}
m_1 <- pfr(Y_1 ~ lf(Z), data = ghcm_sim_data)
m_X <- pffr(X ~ ff(Z), data = ghcm_sim_data, chunk.size = 31000)
test <- ghcm_test(resid(m_X), resid(m_1))
print(test)
```
We get a $p$-value of 0.865.

## Testing independence of functional variables given a functional variable
Finally, we test whether $X \independent W \cond Z$, which could be relevant in creating prediction models for $X$ or $W$ or in simply ascertaining the relationships between the functional variables. We run the test in the following code:
```{r X-W-Z-test}
m_X <- pffr(X ~ ff(Z), data=ghcm_sim_data, chunk.size=31000)
m_W <- pffr(W ~ ff(Z), data=ghcm_sim_data, chunk.size=31000)
test <- ghcm_test(resid(m_X), resid(m_W))
print(test)
```
We get a $p$-value of $0.6304$.

# Bibliography
